Modelling state-switching functional and biologging data with hidden Markov models

### 2020-12-30 Report/Sections/results.tex
read through main text

The HHMM-DFT, which ignores auto-correlation in acceleration, decodes dive types and subdive states similarly to the CarHHMM-DFT, but it is less likely to categorize the behaviour at the beginning and end of dives as subdive state 3 (see Figures 4 and 5 of the supplementary material). In addition, for all three components of each of $\sigma_A^{*(\cdot,1)}$, $\sigma_A^{*(\cdot,2)}$, and $\sigma_A^{*(\cdot,3)}$, the HHMM-DFT produces estimates which are approximately $50$ to $100$ percent larger than those of the CarHHMM-DFT. The estimated uncertainties of the three components of each of $\hat \mu_A^{*(\cdot,1)}$, $\hat \mu_A^{*(\cdot,2)}$, and $\hat \mu_A^{*(\cdot,3)}$ are also less than half of those for the CarHHMM-DFT (see Tables 1 and 2 of the supplementary material). Further, the pseudoresiduals of the HHMM-DFT are noticeably light-tailed and do not follow a standard normal distribution (see Figure 15 of the supplementary material). These findings suggest that the HHMM-DFT is a significantly worse fit to these data than the full CarHHMM-DFT.

@@

Finally, the CarHMM-DFT, which lacks a hierarchical structure, produces fine-scale parameter estimates and state estimates similar to those of the CarHHMM-DFT. However, its lack of hierarchical structure means that it fails to differentiate between short and long dives. This model therefore does not infer the dive-level Markov chain or the relationship between the dive and subdive levels. In particular, the CarHMM-DFT does not indicate that the whale is more likely to be in subdive state 1 when engaged in longer dives compared to shorter dives.

>>>>

The HHMM-DFT, which ignores auto-correlation in acceleration, decodes dive types and subdive states similarly to the CarHHMM-DFT, but it is less likely to categorize the behaviour at the beginning and end of dives as subdive state 3 (see Figures 4 and 5 of the supplementary material). In addition, for all three components of each of $\sigma_A^{*(\cdot,1)}$, $\sigma_A^{*(\cdot,2)}$, and $\sigma_A^{*(\cdot,3)}$, the HHMM-DFT produces estimates which are approximately $50$ to $100$ percent larger than those of the CarHHMM-DFT. The estimated uncertainties of the three components of each of $\hat \mu_A^{*(\cdot,1)}$, $\hat \mu_A^{*(\cdot,2)}$, and $\hat \mu_A^{*(\cdot,3)}$ are also less than half of those for the CarHHMM-DFT (see Tables 1 and 2 of the supplementary material). This suggests that including auto-correlation in the model significantly affects parameter estimates. Further, the pseudoresiduals of the HHMM-DFT are noticeably light-tailed and do not follow a standard normal distribution (see Figure 15 of the supplementary material). These findings suggest that the HHMM-DFT is a significantly worse fit to these data than the full CarHHMM-DFT.

@@

Finally, the CarHMM-DFT, which lacks a hierarchical structure, produces fine-scale parameter estimates and subdive state estimates similar to those of the CarHHMM-DFT. However, its lack of hierarchical structure means that it fails to differentiate between short and long dives. This model therefore does not infer the dive-level Markov chain or the relationship between the dive and subdive levels. In particular, the CarHMM-DFT does not indicate that the whale is more likely to be in subdive state 1 when engaged in longer dives compared to shorter dives.

### 2020-12-30 Report/Sections/simulation.tex
read through main text

\subsection{Data simulation}
\label{subsec:data_simulation}

We generate 500 independent data sets using the CarHHMM-DFT as a generative model. Each data set consists of a sequence of 100 curves which we call a sequence of killer whale dives. Each dive can be one of $N=2$ dive types based on a Markov chain with probability transition matrix
%
$$\Gamma = \begin{pmatrix} 0.79 & 0.21 \\ 0.81 & 0.19 \end{pmatrix}.$$
%
Dive duration is Gamma distributed and the coarse-scale emission parameters are 
%
\[\mu^{(1)} = 25.7s, \enspace \sigma^{(1)} = 9.6s, \enspace \mu^{(2)} = 104.6s, \enspace \sigma^{(2)} = 64.7s.\]
%
After generating the dive durations for all 100 dives in a data set, dive $t$ is broken into a sequence of $\tilde T^*_t = \lfloor Y_t/2 \rfloor$ two-second windows, where the last $Y_t - 2 \tilde T^*_t$ seconds of each simulated dive are ignored. Each two-second segment is assigned one of $N^*=2$ behaviours according to a fine-scale Markov chain $\tilde X^*_t \equiv \left\{\tilde X^*_{t,1}, \ldots, \tilde X^*_{t,\tilde T^*_t} \right\}$ with probability transition matrices
%
\[\Gamma^{*(1)} = \begin{pmatrix} 0.68 & 0.32 \\ 0.05 & 0.95 \end{pmatrix} \quad \text{ and } \quad \Gamma^{*(2)} = \begin{pmatrix} 0.86 & 0.14 \\ 0.15 & 0.85 \end{pmatrix}\]
%
for dive types 1 and 2, respectively.
Instead of generating the raw observations $Y^*_{t,t^*}$, we directly simulate the fine-scale transformed observations $\Z_{t,\tilde t^*} = \left\{\Zone_{t,\tilde t^*}, \Ztwo_{t,\tilde t^*}\right\}$. Recall from Section \ref{subsec:model_selection} that we must specify the mean, standard deviation, and auto-correlation parameters corresponding to $\left\{\Zone_{t,1},\ldots,\Zone_{t,\tilde T_t^*}\right\}$ as well as the mean and standard deviation parameters corresponding to $\left\{\Ztwo_{t,1},\ldots,\Ztwo_{t,\tilde T_t^*}\right\}$. We select the following parameters in line with the results from the case study:
%$\Zone_{t,\tilde t^*}$ and $\Ztwo_{t,\tilde t^*}$ are independent when conditioned on all dive types and subdive states. Each observation of acceleration $\Zone_{t,\tilde t^*}$ is Normally-distributed with explicit auto-correlation while each observation of wiggliness $\Ztwo_{t,\tilde t^*}$ is Gamma-distributed without explicit auto-correlation. In particular, we set the distributions of $\Zone_{t,\tilde t^*}$ and $\Ztwo_{t,\tilde t^*}$ with the following parameters:
%
\begin{gather*}
    \mu_A^{*(\cdot,1)} = 0.0 s, \enspace \sigma_A^{*(\cdot,1)} = 0.034s, \enspace \phi_A^{*(\cdot,1)} = 0.98, \\
    %
    \mu_A^{*(\cdot,2)} = 0.0 s, \enspace \sigma_A^{*(\cdot,2)} = 0.079s, \enspace \phi_A^{*(\cdot,2)} = 0.87, \\
    %
    \mu_W^{*(\cdot,1)} = 23.3, \quad \sigma_W^{*(\cdot,1)} = 13.0, \\
    %
    \mu_W^{*(\cdot,2)} = 301.2, \quad \sigma_W^{*(\cdot,2)} = 330.1.
\end{gather*}
%
It is not possible to uniquely reconstruct the raw accelerometer data $Y^*$ from $\Z$ alone, but we describe one possible mapping from $\Z$ to $Y^*$ in the appendix. Figure 21 of the supplementary material shows one realization of $\Z$ for five dives of one simulated data set along with the the corresponding reconstructed realization of $Y^*$.

@@

We calculate maximum likelihood estimates $\hat \theta$ for all four candidate models for each of the 500 training data sets using the Cedar Compute Canada cluster with 1 CPU and 4 GB of dedicated memory per model. We then generate a corresponding independent test data set consisting of an additional 100 dives from the generative model with the true parameters $\theta$.

If the parameters $\theta$, coarse-scale observations $y$, and fine-scale observations $\z$ are known, then the \textit{forward-backward} algorithm can be used to calculate the hidden state probabilities $\Pr(X_{t} = i|Y=y,\Z=\z)$ and $\Pr(\tilde X^*_{t,t^*} = i^*|Y=y,\Z=\z)$ for all $t$, $t^*$, $i = 1,\ldots,N$, and $i^* = 1,\ldots,N^*$ \citep{Zucchini:2016}. However, since $\theta$ is rarely known in practice, the maximum likelihood estimate $\hat \theta$ is often used instead to estimate the hidden state probabilities. As such, if $X_t$ is the unknown dive type associated with a given model and $x_t$ is the true simulated dive type of dive $t$, then we define the \textit{average dive decoding accuracy} as the average value of the estimates of $\Pr(X_{t} = x_t|Y=y,\Z=\z)$ across all simulated data sets and dives. Analogously, if $\tilde X^*_{t,\tilde t^*}$ is the unknown subdive state associated with a particular model and $\tilde x^*_{t,\tilde t^*}$ is the true simulated subdive state for dive $t$ and window $\tilde t^*$, then we define the \textit{average subdive decoding accuracy} as the average value of the estimates of $\Pr(\tilde X^*_{t,t^*} = \tilde x^*_{t,\tilde t^*}|Y=y,\Z=\z)$ across all simulated data sets, dives, and windows. We find both the average dive decoding accuracy and the average subdive decoding accuracy using the maximum likelihood estimates $\hat \theta$ for each model and training set, the observations $y$ and $\z$ of each test set, and the forward-backward algorithm.

@@

The full CarHHMM-DFT is the best performing model of the four candidates since it is the generating model. Its average dive decoding accuracy and the average subdive decoding accuracy are both greater than 0.9. All parameter estimates of fine-scale mean values and probability transition matrices, $\hat \mu^*$, $\hat \Gamma^*$ and $\hat \Gamma$, respectively, have statistically insignificant biases. In addition, the biases of $\hat \sigma$, $\hat \phi$, and $\hat \mu$ are either as small or smaller than the other three candidate models. The empirical standard errors of all parameter estimates ($\hat \theta$, $\hat \Gamma$, $\hat \theta^*$, $\hat \Gamma^*$) are well-approximated by the inverse of the observed Fisher information matrix, although the estimated standard errors tend to be slightly smaller than the empirical standard errors. This small bias is especially noticeable for parameters associated with the wiggliness $\Ztwo_{t,\tilde t^*}$, where the empirical standard error can be up to twice as large as the estimated standard error. See Tables 5 through 9 of the supplementary material for detailed results.

@@

Figures \ref{fig:acc_coarse} and \ref{fig:acc_fine} display one simulated data set as well as the decoded dive types and subdive states associated with each model. The CarHHMM-DFT and CarHMM-DFT produce similar estimates of subdive state while the HHMM-DFT is slightly more likely to predict that a given window corresponds to subdive state 2. As expected, the CarHHMM is the least accurate model when predicting subdive state. The CarHHMM-DFT and HHMM-DFT yield similar estimates of dive type while the CarHHMM is less accurate and even misclassifies the second dive. The CarHMM-DFT does not estimate dive type.

>>>>

\subsection{Simulation procedure}
\label{subsec:data_simulation}

We generate 500 independent data sets using the CarHHMM-DFT as a generative model. Each data set consists of a sequence of 100 curves which we call a sequence of killer whale dives. Each dive can be one of $N=2$ dive types based on a Markov chain with probability transition matrix
%
$$\Gamma = \begin{pmatrix} 0.79 & 0.21 \\ 0.81 & 0.19 \end{pmatrix}.$$
%
Dive duration is Gamma distributed and the coarse-scale emission parameters are 
%
\[\mu^{(1)} = 25.7s, \enspace \sigma^{(1)} = 9.6s, \enspace \mu^{(2)} = 104.6s, \enspace \sigma^{(2)} = 64.7s.\]
%
After generating the dive durations for all 100 dives in a data set, dive $t$ is broken into a sequence of $\tilde T^*_t = \lfloor Y_t/2 \rfloor$ two-second windows, where the last $Y_t - 2 \tilde T^*_t$ seconds of each simulated dive are ignored. Each two-second segment is assigned one of $N^*=2$ behaviours according to a fine-scale Markov chain $\tilde X^*_t \equiv \left\{\tilde X^*_{t,1}, \ldots, \tilde X^*_{t,\tilde T^*_t} \right\}$ with probability transition matrices
%
\[\Gamma^{*(1)} = \begin{pmatrix} 0.68 & 0.32 \\ 0.05 & 0.95 \end{pmatrix} \quad \text{ and } \quad \Gamma^{*(2)} = \begin{pmatrix} 0.86 & 0.14 \\ 0.15 & 0.85 \end{pmatrix}\]
%
for dive types 1 and 2, respectively.
Instead of generating the raw observations $Y^*_{t,t^*}$, we directly simulate the fine-scale transformed observations $\Z_{t,\tilde t^*} = \left\{\Zone_{t,\tilde t^*}, \Ztwo_{t,\tilde t^*}\right\}$. Recall from Section \ref{subsec:model_selection} that we must specify the mean, standard deviation, and auto-correlation parameters corresponding to $\left\{\Zone_{t,1},\ldots,\Zone_{t,\tilde T_t^*}\right\}$ as well as the mean and standard deviation parameters corresponding to $\left\{\Ztwo_{t,1},\ldots,\Ztwo_{t,\tilde T_t^*}\right\}$. We select the following parameters in line with the results from the case study:
%$\Zone_{t,\tilde t^*}$ and $\Ztwo_{t,\tilde t^*}$ are independent when conditioned on all dive types and subdive states. Each observation of acceleration $\Zone_{t,\tilde t^*}$ is Normally-distributed with explicit auto-correlation while each observation of wiggliness $\Ztwo_{t,\tilde t^*}$ is Gamma-distributed without explicit auto-correlation. In particular, we set the distributions of $\Zone_{t,\tilde t^*}$ and $\Ztwo_{t,\tilde t^*}$ with the following parameters:
%
\begin{gather*}
    \mu_A^{*(\cdot,1)} = 0.0 s, \enspace \sigma_A^{*(\cdot,1)} = 0.034s, \enspace \phi_A^{*(\cdot,1)} = 0.98, \\
    %
    \mu_A^{*(\cdot,2)} = 0.0 s, \enspace \sigma_A^{*(\cdot,2)} = 0.079s, \enspace \phi_A^{*(\cdot,2)} = 0.87, \\
    %
    \mu_W^{*(\cdot,1)} = 23.3, \quad \sigma_W^{*(\cdot,1)} = 13.0, \\
    %
    \mu_W^{*(\cdot,2)} = 301.2, \quad \sigma_W^{*(\cdot,2)} = 330.1.
\end{gather*}
%
It is not possible to uniquely reconstruct the raw accelerometer data $Y^*$ from $\Z$ alone, but we describe one possible mapping from $\Z$ to $Y^*$ in the appendix. Figure 21 of the supplementary material shows one realization of $\Z$ for five dives of one simulated data set along with the corresponding reconstructed realization of $Y^*$.

@@

We calculate maximum likelihood estimates $\{\hat \theta, \hat \Gamma, \hat \theta^*, \hat \Gamma^*\}$ for all four candidate models for each of the 500 training data sets using the Cedar Compute Canada cluster with 1 CPU and 4 GB of dedicated memory per model. We then generate a corresponding independent test data set consisting of an additional 100 dives from the generative model with the true parameters $\{\theta, \Gamma, \theta^*, \Gamma^*\}$.

If the parameters, coarse-scale observations $y$, and fine-scale observations $\z$ are known, then the \textit{forward-backward} algorithm can be used to calculate the hidden state probabilities $\Pr(X_{t} = i|Y=y,\Z=\z)$ and $\Pr(\tilde X^*_{t,t^*} = i^*|Y=y,\Z=\z)$ for all $t$, $t^*$, $i = 1,\ldots,N$, and $i^* = 1,\ldots,N^*$ \citep{Zucchini:2016}. However, since the true parameters rarely known in practice, the maximum likelihood estimates are often used instead to estimate the hidden state probabilities. As such, if $X_t$ is the unknown dive type associated with a given model and $x_t$ is the true simulated dive type of dive $t$, then we define the \textit{average dive decoding accuracy} as the average value of the estimates of $\Pr(X_{t} = x_t|Y=y,\Z=\z)$ across all simulated data sets and dives. Analogously, if $\tilde X^*_{t,\tilde t^*}$ is the unknown subdive state associated with a particular model and $\tilde x^*_{t,\tilde t^*}$ is the true simulated subdive state for dive $t$ and window $\tilde t^*$, then we define the \textit{average subdive decoding accuracy} as the average value of the estimates of $\Pr(\tilde X^*_{t,t^*} = \tilde x^*_{t,\tilde t^*}|Y=y,\Z=\z)$ across all simulated data sets, dives, and windows. We find both the average dive decoding accuracy and the average subdive decoding accuracy using the maximum likelihood estimates for each model and training set, the observations $y$ and $\z$ of each test set, and the forward-backward algorithm.

@@

The full CarHHMM-DFT is the best performing model of the four candidates since it is the generating model. Its average dive decoding accuracy and average subdive decoding accuracy are both greater than 0.9. All parameter estimates of fine-scale mean values and probability transition matrices, $\hat \mu^*$, $\hat \Gamma^*$ and $\hat \Gamma$, respectively, have statistically insignificant biases. In addition, the biases of $\hat \sigma$, $\hat \phi$, and $\hat \mu$ are either as small or smaller than the other three candidate models. The empirical standard errors of all parameter estimates ($\hat \theta$, $\hat \Gamma$, $\hat \theta^*$, $\hat \Gamma^*$) are well-approximated by the inverse of the observed Fisher information matrix, although the estimated standard errors tend to be slightly smaller than the empirical standard errors. This underestimation is especially noticeable for parameters associated with the wiggliness $\Ztwo_{t,\tilde t^*}$, where the empirical standard error can be up to twice as large as the estimated standard error. See Tables 5 through 9 of the supplementary material for detailed results.

@@

Figures \ref{fig:acc_coarse} and \ref{fig:acc_fine} display one simulated data set as well as the decoded dive types and subdive states associated with each model. The CarHHMM-DFT and CarHMM-DFT produce similar estimates of subdive state while the HHMM-DFT is slightly more likely to predict that a given window corresponds to subdive state 2. As expected, the CarHHMM is the least accurate model when predicting subdive state. The CarHHMM-DFT and HHMM-DFT yield similar estimates of dive type while the CarHHMM is less accurate and even misclassifies the dive type of the second dive. The CarHMM-DFT does not estimate dive type.
